Example Kitsap County Pedestrian Facilities Evaluation Report

### 2022-10-31 Draft-Ped-Metrics.Rmd
many updates not tracked

---
title: "DRAFT W GRAVEL! Example Kitsap County Pedestrian Facilities Metrics/Evaluation Report"
date: "`r Sys.Date()`"
bibliography: metrics.bib
output: 
  html_document:
    number_sections: true
    toc: true
    toc_float: true
    toc_depth: 3
    fig_caption: yes
---

@@

nice_comma <- function(x, dec = 1){
  format(round(as.numeric(x), dec), digits = NULL, nsmall=dec, big.mark=",") 
}
'%ni%' <- function(x,y)!('%in%'(x,y))

@@

library(kableExtra)
library(tidyverse)
library(sf)
```

@@

source(paste0(this_dir,'functions/getCleanRoadCenterLine.R'))
library(tidyverse)
library(sf)

outline <- sf::read_sf(paste0(this_dir,"data/outline"))
outline$plain = 1
kitsap <- ggplot() + 
  geom_sf(data = outline["plain"], alpha = 0.5) + 
  theme_void()

cities <- sf::read_sf(paste0(this_dir,"data/cities"))

# try to make a kitsap without cities
county_no_cities <- st_difference(outline, st_union(cities))

uga <- sf::read_sf(paste0(this_dir,"data/uga"))
uga$isuga = 1
lamird <- sf::read_sf(paste0(this_dir,"data/lamird"))
lamird$islamird = 1
community <- sf::read_sf(paste0(this_dir,"data/community"))
schools <- sf::read_sf(paste0(this_dir,"data/schools"))

# start plotting roads
# want always to see same color for same type of road, so
#['#e41a1c','#377eb8','#4daf4a','#984ea3','#ff7f00','#ffff33','#a65628','#f781bf']
# NOTE: Also make this a factor so that it is sorted properly!!
road_col <- scale_color_manual(values = c("Freeway / Expressway" = "#e41a1c",
                                          "Principal Arterial" = "#377eb8",
                                          "Minor Arterial" = '#4daf4a',
                                          "Major Collector" = "#984ea3",
                                          "Minor Collector" ='#ff7f00',
                                          "Local Access" = '#a65628',
                                          "Rural Local Access (e.g. easement or other type)" = '#f781bf',
                                          "Urban Local Access (e.g. easement or other type)" = '#ffff33'
                                          ))
clean_roadcl$road_class <- factor(clean_roadcl$road_class, levels = c("Freeway / Expressway",
                                                                      "Principal Arterial",
                                                                      "Minor Arterial",
                                                                      "Major Collector",
                                                                      "Minor Collector",
                                                                      "Local Access",
                                                                      "Rural Local Access (e.g. easement or other type)",
                                                                      "Urban Local Access (e.g. easement or other type)"))

# exclude Local Access EASEMENTS & Highway/Freeway
road_df <-st_as_sf(clean_roadcl) %>%
  filter(FCLASS %ni% c(100,110,1,2), RD_LOG_ID > 0) %>%
  mutate(LENGTH = as.integer(st_length(.)))

# need to make sure that this is right

ncdf <- st_intersection(road_df, st_buffer(county_no_cities,0)) %>%
  mutate(LENGTH = as.integer(st_length(.)))

@@

```{r swshdata, cache = TRUE}
library(tidyverse)
library(sf)
shoulders <- sf::read_sf("./data/KitsapShapefiles03_04_2022/SHOULDERS")
sidewalks <- sf::read_sf("./data/KitsapShapefiles03_04_2022/SIDEWALKS")

sidewalks_nc <- st_intersection(sidewalks, st_buffer(county_no_cities,0)) %>%
  st_as_sf(.) %>%
  filter(WIDTH > 0, LENG_FT > 0) %>%
  mutate(LENGTH = as.integer(st_length(.)),
         TYPE = paste0("Sidewalk: ",WIDTH, " ft"),
         RD_LOG_ID = as.integer(ROADLOGID),
         SEGMENT_ID = as.integer(SEG_ID)) %>%
  select(RD_LOG_ID,SEGMENT_ID,BMP,EMP,SIDE_OF_RO, WIDTH, LENGTH, TYPE)

# SHLDR_SURF 
# ACP - paved
# BST - bituminous surface (chip seal), like paved
# GRV - gravel
# PAP - porous asphalt
shoulders_nc <- st_intersection(shoulders, st_buffer(county_no_cities,0)) %>%
  st_as_sf(.) %>%
  filter(SHLDR_WIDT > 0, 
         SHLDR_SURF %in% c("ACP", "BST", "PAP", "GRV"), # these are the "paved" ones
         DISTANCE_F > 0) %>%
  mutate(LENGTH = as.integer(st_length(.)),
         TYPE = paste0("Shoulder: ", SHLDR_WIDT," ft"),
         RD_LOG_ID = as.integer(ROAD_LOG_I),
         SEGMENT_ID = as.integer(SEG_ID)) %>%
  select(RD_LOG_ID,SEGMENT_ID,BMP,EMP,SIDE_OF_RO, WIDTH = SHLDR_WIDT, LENGTH, TYPE)
nc_sw_sh <- bind_rows(sidewalks_nc, shoulders_nc)

# NOTE: Also make this a factor so that it is sorted properly!!
sw_sh_col <- scale_color_manual(values = c("Sidewalk: 3 ft"="#E41A1C",
                                           "Sidewalk: 4 ft"="#874F6F",
                                           "Sidewalk: 5 ft"="#3881B0",
                                           "Sidewalk: 8 ft"="#449B75",
                                           "Shoulder: 1 ft"="#56A255",
                                           "Shoulder: 2 ft"="#7E6E85",
                                           "Shoulder: 3 ft"="#AC5782",
                                           "Shoulder: 4 ft"="#E3712B", 
                                           "Shoulder: 5 ft"="#FFA10D",
                                           "Shoulder: 6 ft"="#FFE528",
                                           "Shoulder: 7 ft"="#E1C62F",
                                           "Shoulder: 8 ft"="#B16C29",
                                           "Shoulder: 10 ft"="#C66764",
                                           "Shoulder: 12 ft"="#F17EB4",
                                           "Shoulder: 15 ft"="#CB8CAD",
                                           "Shoulder: 17 ft"="#999999"))
nc_sw_sh$TYPE <- factor(nc_sw_sh$TYPE, levels = c("Sidewalk: 3 ft",
                                                  "Sidewalk: 4 ft",
                                                  "Sidewalk: 5 ft",
                                                  "Sidewalk: 8 ft" ,
                                                  "Shoulder: 1 ft",
                                                  "Shoulder: 2 ft",
                                                  "Shoulder: 3 ft",
                                                  "Shoulder: 4 ft", 
                                                  "Shoulder: 5 ft",
                                                  "Shoulder: 6 ft",
                                                  "Shoulder: 7 ft",
                                                  "Shoulder: 8 ft",
                                                  "Shoulder: 10 ft",
                                                  "Shoulder: 12 ft",
                                                  "Shoulder: 15 ft",
                                                  "Shoulder: 17 ft"))
```

```{r gwfunction}
getWalkingPolygonInCounty <- function(locdf, idcol) {
  # finds the items in locdf that are in the target area
  # then finds the polygon around them
  # then limits that again to the area in the target (because the polygon may go into a city)
  if (!is.na(st_crs(locdf)$epsg)) { # the locdf has a epsg and cnc is NA
    cnc <- st_transform(county_no_cities, st_crs(locdf)$epsg)
    fdf <- st_intersection(locdf, st_buffer(cnc,0)) 
  } else {
    fdf <- st_intersection(locdf, st_buffer(county_no_cities,0)) 
  }
  fdf2 <- mapboxapi::mb_isochrone(fdf,
                                  time = 30, "walking", id_column=idcol)
  cnc <- st_transform(county_no_cities, st_crs(fdf2)$epsg)
  st_intersection(fdf2, st_buffer(cnc,0)) 
}
```

```{r smallerdata}
# this section sets up the data that will be used for subdivisions of the county
# get any data not set up yet
commdists <- sf::read_sf(paste0(this_dir,"data/commdist"))
commdists <- st_intersection(commdists, st_buffer(county_no_cities,0)) 
commdists <- st_as_sf(commdists) %>% 
  mutate(NAME = ifelse(DISTRICT == "1", "1 North",
                ifelse(DISTRICT == "2", "2 South", "3 Central")))
the_commdists <- unique(st_set_geometry(commdists, NULL)$NAME)
loop_commdists <- st_as_sf(commdists) %>% mutate(tomatch = NAME)

uga_no_city <- st_as_sf(uga) %>%
  filter(substring(GMA_JURISD, 1, 4) != "City")
the_ugas <- unique(st_set_geometry(uga_no_city, NULL)$GMA_JURISD)
loop_ugas <- st_as_sf(uga_no_city) %>% mutate(tomatch = GMA_JURISD)

the_lamirds <- unique(st_set_geometry(lamird, NULL)$PLAN_AREA)
loop_lamirds <- st_as_sf(lamird) %>% mutate(tomatch = PLAN_AREA)

pub_schools <- schools %>%
  filter(TYPE == "PUBLIC")
pub_schools2 <- getWalkingPolygonInCounty(pub_schools, "NAME")
the_schools <- unique(st_set_geometry(pub_schools2, NULL)$id)
loop_schools <- st_as_sf(pub_schools2) %>% mutate(tomatch = id)

libraries <- community %>%
  filter(TYPE == "PUBLIC LIBRARY")
libraries2 <- getWalkingPolygonInCounty(libraries, "NAME")
the_libraries <- unique(st_set_geometry(libraries2, NULL)$id)
loop_libraries <- st_as_sf(libraries2) %>% mutate(tomatch = id)

transit_cp <- tribble(
  ~NAME, ~ADDRESS,
  "Bremerton Transportation Center (Ferry Terminal)","10 Washington Ave, Bremerton 98337",
  "Bainbridge Island Ferry Terminal","270 Olympic Drive SE, Bainbridge 98110",
  "Wheaton Way Transit Center","3915 Wheaton Way, Bremerton 98310",
  "West Bremerton Transit Center","540 Bruenn Ave, Bremerton 98312",
  "Silverdale Transit Center","Greaves Way & Kitsap Mall Blvd, Silverdale 98383",
  "North Viking Transit Center","21992 Viking Ave NW, Poulsbo 98370",
  "Kingston Ferry Terminal","11264 State Route 104, Kingston 98346",
  "Port Orchard Ferry Dock","73 Sidney Ave, Port Orchard 98366",
  "Southworth Ferry Terminal","11564 SE State Hwy. 160, Southworth 98386",
  "Annapolis Ferry Dock","1076 Bay Street, Port Orchard 98366",
  "Hwy. 305 & Suquamish Way","16003 WA-305, Poulsbo, Washington 98370",
  "McWilliams Park & Ride","1601 NE McWilliams Rd, Bremerton, Washington 98311",
  "Suquamish Park & Ride","18829 Division Ave NE, Suquamish, Washington 98392",
  "Port Orchard Wal-Mart","3497 Bethel Rd SE, Port Orchard 98366",
  "George's Corner Park & Ride","27618 Hansville Rd NE, Kingston 98346",
  "Gateway Fellowship Park & Ride","18901 8th Ave NE, Poulsbo 98370",
  "Miller Bay & Indianola","23404 Miller Bay Rd NE Poulsbo, Washington 98370",
  "St. Gabriel's Church","1150 Mitchell Ave SE, Port Orchard 98366"
)
# geocode the addresses into geocode = c(lon, lat)
transit_cp <- transit_cp %>% #tmaptools::geocode_OSM(transit_cp$ADDRESS)
  rowwise() %>%
  mutate(geocode = list(set_names(mapboxapi::mb_geocode(ADDRESS),
                        c("lon","lat")))) %>%
  unnest_wider(geocode)

# convert the lat lon into a point geometry
transits <- st_as_sf(transit_cp, 
                      coords = c(x = "lon", y = "lat"), 
                      crs = 4326) 
transits2 <- getWalkingPolygonInCounty(transits, "NAME")
the_transits <- unique(st_set_geometry(transits2, NULL)$id)
loop_transits <- st_as_sf(transits2) %>% mutate(tomatch = id)

# combine data into tibble for this report
#These regions and pedestrian generators are:
kds <- "Kitsap County GIS"
kt <- "Kitsap Transit Website"
wd <- "Walking distance polygons from Mapbox"
breakdowns <- tribble(
  ~Name,~pointdf,~polydf,~idcolpoly,~items,~data_source,
  "Commissioner Districts",NA,"commdists","NAME","the_commdists",kds,
  "UGAs",NA,"uga_no_city","GMA_JURISD","the_ugas",kds,
  "LAMIRDs",NA,"lamird","PLAN_AREA","the_lamirds",kds,
  "Public Schools","pub_schools","pub_schools2","id","the_schools",paste(kds,wd),
  "Public Libraries","libraries","libraries2","id","the_libraries",paste(kds,wd),
  "Transit Centers","transits","transits2","id","the_transits",paste(kt,wd),
  )

@@

```{r kitsapmap, cache = TRUE}

#| fig.align = "center"
#| fig.width = 12
#| fig.height = 8
#| out.width = '100%'

@@

kitsap_all_roads <- kitsap +
  geom_sf(data = clean_roadcl, aes(color = road_class)) + 
  road_col +
  theme_void() +
  labs(title = "Roads in Kitsap County",
       subtitle = paste0(nice_comma(sum(st_length(clean_roadcl))/5280),
                         " centerline miles"),
       #caption = "Data source: Kitsap GIS",
       color = "")

kitsap_roads_no_cities <- kitsap + 
  geom_sf(data = cities["NAME"], 
          aes(fill="NAME"),
          fill="#808080",
          alpha = 0.5,
          show.legend = FALSE) +
  geom_sf(data = ncdf, aes(color = road_class), show.legend = FALSE) + 
  road_col +
  theme_void() +
  labs(title = "No Cities, Easements, & Highways",
       subtitle = paste0(nice_comma(sum(st_length(ncdf))/5280),
                         " centerline miles"),
       caption = "Data source: Kitsap GIS")

library(patchwork)

kitsap_all_roads + kitsap_roads_no_cities +
  plot_layout(guides = 'collect') & theme(legend.position = 'bottom')

detach("package:patchwork", unload=TRUE)
```

@@

These are the roads that will be evaluated.
```{r}
library(tidyverse)
library(sf)

groupcl <- ncdf %>%  
  st_set_geometry(., NULL) %>% 
  group_by(road_class) %>% 
  summarise(roads = nice_comma(n_distinct(FULL_NAME),0), 
            segments = nice_comma(n(),0), 
            length = nice_comma(sum(LENGTH, na.rm=TRUE)/5280, 1),
            .groups = "drop") 
df <- ncdf %>% #clean_roadcl %>%  
  st_set_geometry(., NULL) %>% 
  group_by(road_class = "Total") %>% 
  summarise(roads = nice_comma(n_distinct(FULL_NAME),0), 
            segments = nice_comma(n(),0), 
            length = nice_comma(sum(LENGTH, na.rm=TRUE)/5280, 1),
            .groups = "drop") 
groupcl <- bind_rows(groupcl, df)

colnames(groupcl) <- c("Road Classification", 
                       "Count of Roads", "Count of Segments", 
                       "Length in Miles") 
knitr::kable(groupcl, 
             caption = "Roads by Class in Kitsap County, excluding cities, highways, and easements")  %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

@@

```{r}
df <- sidewalks_nc %>%  
  st_set_geometry(., NULL) %>%
  mutate(`Width in Feet` = paste(WIDTH, "ft")) %>%
  group_by(WIDTH, `Width in Feet` ) %>% # road_type) %>%
  summarise(miles = nice_comma(sum(LENGTH, na.rm=TRUE)/5280,1),
            .groups = "drop") %>%
  select(-WIDTH)
df2 <- sidewalks_nc %>%  
  st_set_geometry(., NULL) %>%
  mutate(`Width in Feet` = "Total", WIDTH = 100) %>%
  group_by(WIDTH, `Width in Feet` ) %>% # road_type) %>%
  summarise(miles = nice_comma(sum(LENGTH, na.rm=TRUE)/5280,1),
            .groups = "drop") %>%
  select(-WIDTH)

df <- bind_rows(df, df2)

knitr::kable(df, 
             caption = "Miles of Sidewalks by Width of Sidewalk (single side)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

```{r}
df <- shoulders_nc %>%  
  st_set_geometry(., NULL) %>%
  mutate(`Width in Feet` = paste(WIDTH, "ft")) %>%
  group_by(WIDTH, `Width in Feet` ) %>% # road_type) %>%
  summarise(miles = nice_comma(sum(LENGTH, na.rm=TRUE)/5280,1),
            .groups = "drop") %>%
  select(-WIDTH)
df2 <- shoulders_nc %>%  
  st_set_geometry(., NULL) %>%
  mutate(`Width in Feet` = "Total", WIDTH = 100) %>%
  group_by(WIDTH, `Width in Feet` ) %>% # road_type) %>%
  summarise(miles = nice_comma(sum(LENGTH, na.rm=TRUE)/5280,1),
            .groups = "drop") %>%
  select(-WIDTH)

df <- bind_rows(df, df2)

knitr::kable(df, 
             caption = "Miles of Paved Shoulder by Width of Shoulder (single side)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

@@

In order to do this evaluation, we must combine the roads data (which has information on the class and setting of roads) with the sidewalk and shoulder data (with information about the pedestrian facilities). However, the most obvious methods of combination fail.

#### Matching on Road Log Id is good but insufficient

All of the data sets appear to use the same set of road log IDs from public works. Unfortunately, knowing that there is a shoulder or sidewalk somewhere on a road does not get to how much of that road has the sidewalk or shoulder.^[As roads sometimes change names and road log IDs, there are also some breaks between the road data and sidewalk/shoulder data. This was noted for Madrone Ave in Manchester - given the same road log ID as Main St when the road bends.]

This diagnostic and ugly plot is intended to show the coverage of road log id across the sidewalk and shoulder data sets compared to the raod data. While the sidewalk and shoulder data are more sparse, the range is consistent.

```{r}
sw_plot <- ggplot(st_set_geometry(nc_sw_sh,NULL), 
                  aes(x = as.numeric(row.names(nc_sw_sh)), y = RD_LOG_ID)) + 
  geom_point() +
  labs(title = "Sidewalk and Shoulder Data",
       x = "") +
  expand_limits(y = c(0,100000))
road_plot <- ggplot(st_set_geometry(ncdf,NULL), 
                  aes(x = as.numeric(row.names(ncdf)), y = RD_LOG_ID)) + 
  geom_point() +
  labs(title = "Roads Data",
       x = "")

library(patchwork)

sw_plot + road_plot

detach("package:patchwork", unload=TRUE)
rm(sw_plot)
rm(road_plot)
```

```{r}
match_road_id <- st_set_geometry(sidewalks,NULL) %>%
  mutate(RD_LOG_ID = as.integer(ROADLOGID)) %>%
  left_join(select(st_set_geometry(clean_roadcl,NULL),RD_LOG_ID,FULL_NAME)) %>%
  group_by(RD_LOG_ID, ROADNAME, SIDE_OF_RO, WIDTH) %>%
  summarise(n = sum(!is.na(FULL_NAME)), .groups = "drop") %>% 
  filter(n == 0, WIDTH > 0)
```

The most obvious smaller match is the segment id. However, 
**NOTE: We cannot actually use the SEGMENT ID in this draft report because it does not have the same meaning in the sidewalk and shoulder data as it has in the road data. The following plots are diagnostic but show that the sidewalks and shoulders data do not have segment ids covering the same spectrum as the road data.

```{r}
sw_plot <- ggplot(st_set_geometry(nc_sw_sh,NULL), 
                  aes(x = as.numeric(row.names(nc_sw_sh)), y = SEGMENT_ID)) + 
  geom_point() +
  labs(title = "Sidewalk and Shoulder Data",
       x = "")+
  expand_limits(y = c(0,20000))
road_plot <- ggplot(st_set_geometry(ncdf,NULL), 
                  aes(x = as.numeric(row.names(ncdf)), y = SEGMENT_ID)) + 
  geom_point() +
  labs(title = "Roads Data",
       x = "")

library(patchwork)

sw_plot + road_plot

detach("package:patchwork", unload=TRUE)
rm(sw_plot)
rm(road_plot)
```

The segment ids are clearly not from the same space, but the road log ids are consistent.^[Mostly.] However, we cannot combine on road log id alone - because the roads change character and the type of pedestrian facilities over space. We combine instead on geometry.

For now, we are making a space around a sidewalk or shoulder geometry, a buffer, to capture the nearby area where the road would be, and seeing if there are any road segments that intersect. If the road segment intersects and has the same road log id, we keep it.

```{r combine-rd-sw-sh}
# combine the sidewalk and shoulder and roads data
x <- st_as_sf(nc_sw_sh) %>% mutate(LENGTH = as.integer(LENGTH)) %>% filter(RD_LOG_ID == 49435 & LENGTH > 600)
x$RD_LOG_ID <- 43809
comb <- st_intersection(select(ncdf, RD_LOG_ID_R = RD_LOG_ID, 
                                     SEGMENT_ID_R = SEGMENT_ID,
                                     FULL_NAME_R = FULL_NAME, 
                                     LENGTH_R = LENGTH,
                                     road_class, RUCODE), 
                             st_buffer(bind_rows(nc_sw_sh, x), 50, endCapStyle="FLAT")) %>% 
  filter(RD_LOG_ID == RD_LOG_ID_R) %>%
  distinct(RD_LOG_ID_R, SEGMENT_ID_R, RD_LOG_ID, SEGMENT_ID, SIDE_OF_RO, .keep_all = TRUE) %>%
  mutate(distance = as.integer(st_length(.)))

@@

Road segments are the basis for evaluation because they are the smallest unit in common across the data and can be grouped together to form trips. Actually matching up the data for roads, shoulders, and sidewalks will prove to be a more complicated task than it appears at first blush. As noted in the [Data section](#datasection), the Segment IDs in the shoulder and sidewalk data do not cover the range of segments as in the road data. Here, we join based on the geometries of the roads and sidewalks/shoulders.^[We intersect the road data with a buffered geometry of the sidewalk or shoulder and then remove any matching roads that do not have the same road log id.]

@@

```{r tbl_standards}
# Road Standards Tables 3.3 and 3.4
library(tidyverse)
library(kableExtra)
road_standards_given <- tribble(
  ~setting, ~class, ~characteristics, ~sidewalk, ~shoulder, 
  "Urban", "Local Road", NA, 5, NA,
  "Urban", "Local Sub-collector", NA, 5, NA,
  "Urban", "Arterial", NA, 6, NA,
  "Urban", "Collector", NA, 6, NA,
  "Rural", "Local Road", NA, NA, 3,
  "Rural", "Local Sub-collector", NA, NA, 4,
  "Rural", "Collector", "ADT 400-750", NA, 3,
  "Rural", "Collector", "ADT 751-1000", NA, 4,
  "Rural", "Collector", "DHV 100-200", NA, 6,
  "Rural", "Collector", "DHV > 200", NA, 8,
  "Rural", "Minor Arterial", "DHV < 100", NA, 4,
  "Rural", "Minor Arterial", "DHV 100-200", NA, 6,
  "Rural", "Minor Arterial", "DHV > 200", NA, 8,
  "Rural", "Principal Arterial", "DHV <200", NA, 6,
  "Rural", "Principal Arterial", "DHV > 200", NA, 8,
)

knitr::kable(road_standards_given,
             caption = "Summary of Tables 3.3 and 3.4 in Kitsap County Road Standards [@kc_road_standards]")  %>%
   footnote(general = "Within the road standards, the minimum shoulder width may be reduced to the minimum required by AASHTO, which would be 4-8 feet, depending on the ADT. Note: sub-collector, ADT, and DHV are not defined in the publicly available road data.") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

Because of all of the items that are not available in the road data at this time, this draft makes some assumptions to allow for an evaluation. When the information is known, a more accurate evaluation can be made.^[Roads with unknown setting - RUCODE is 0, we match them to the Urban road standards. This may be incorrect.]

```{r tbl_standards_simple}
# Road Standards Tables 3.3 and 3.4 translated for data we have
library(tidyverse)
library(kableExtra)
road_standards <- tribble(
  ~setting, ~class, ~sidewalk, ~shoulder,
  "Urban", "Local Access", 5, NA,
  "Urban", "Minor Collector",6, NA,
  "Urban", "Major Collector",6, NA,
  "Urban", "Minor Arterial", 6, NA,
  "Urban", "Principal Arterial", 6, NA,
  "Rural", "Local Access", NA, 3,
  "Rural", "Minor Collector",NA, 4,
  "Rural", "Major Collector",NA, 6,
  "Rural", "Minor Arterial", NA, 4,
  "Rural", "Principal Arterial", NA, 6,
)

knitr::kable(road_standards,
             caption = "Simplified Assumptions to Allow for Evaluation")  %>%
   footnote(general = "This is only a simplification to allow for an evaluation without knowledge of ADT or DHV on the roads under evaluation") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

@@

- 5 if it has a sidewalk or shoulder of adequate size (based on road classification & standards) or a sidewalk is present where a shoulder is the standard
- 3 if the sidewalk or shoulder is present and is minimum AASHTO requirement (like a 4 ft shoulder)
- 1 if the sidewalk or shoulder is present and is narrower than AASHTO (**NOTE to subcommittee - do we think we should use a 1 if there is a shoulder when there should be a sidewalk?**)
- 0 if none present or no data

@@

```{r}
example_tbl <- tribble(
  ~Segment, ~"Segment Score", ~"Segment Centerline Linear Feet",
  "A", 6, 100,
  "B", 0, 10,
  "C",10,50,
  "D",8,200,
  "Total",7.5 ,360,
)

knitr::kable(example_tbl) %>%
kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
              full_width = F, 
              position = "float_left")
```

@@

```{r}
score_interpretation <- tribble(
  ~score_at_least, ~interpretation,
  10, "Good",
  5, "Adequate",
  3, "Minimum",
  1, "Poor",
  0.001, "Very Poor",
  0, "Nothing"
)

getScoreInterpretation <- function(x){
  first(score_interpretation$interpretation[
        score_interpretation$score_at_least <= x]
    )
}

df <- score_interpretation %>%
  mutate(`Score At Least` = ifelse(score_at_least == 0.001, "> 0", 
                                   nice_comma(score_at_least,0))) %>%
  select(`Score At Least`, Interpretation = interpretation)

knitr::kable(df,
             caption = "Interpretation of Scores for Segments and for Areas")  %>%
   footnote(general = "For a segment, adequate means one side is good or both sides meet AASHTO minimums and minimal means one side meets AASHTO minimums") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                full_width = F)
```

@@

```{r compare_standards}
# take the road cl data for the roads under evaluation
# remove geometry
# merge with standards to get requirement based on setting (RUCODE) and class (road_class)
# merge with sidewalk and shoulder data to get what exists

road_standards <- road_standards %>%
  rename(standard_sidewalk = sidewalk, 
         standard_shoulder = shoulder, 
         road_class = class)

getSideScore <- function(standard_sidewalk, standard_shoulder,
                         sidewalk, shoulder) {
  ifelse(is.na(standard_sidewalk), # dealing with shoulder
         case_when(
           !is.na(sidewalk) ~ 5, # there is a sidewalk present when a shoulder is the standard
           shoulder >= standard_shoulder ~ 5,
           shoulder < standard_shoulder & shoulder > 4 ~ 3,
           shoulder < standard_shoulder & shoulder > 0 ~ 1,
           TRUE ~ 0
         ),
         case_when(
           sidewalk >= standard_sidewalk ~ 5,
           sidewalk < standard_sidewalk & sidewalk > 4 ~ 3,
           sidewalk < standard_sidewalk & sidewalk > 0 ~ 1,
           TRUE ~ 0
         )
  )
}

comb_stand <- comb %>% 
  st_set_geometry(., NULL) %>% 
  mutate(setting = ifelse(RUCODE == 1, "Rural", 
                          ifelse(RUCODE == 2, "Urban", "Unknown")),
         ped_type = sub("\\:.*", "", TYPE),
         nn = paste(sub("paved ","",tolower(ped_type)), tolower(SIDE_OF_RO),sep ="_")) %>%
  pivot_wider(id_cols = c(-RD_LOG_ID:-TYPE,-ped_type), 
              names_from = nn, 
              values_from = WIDTH,
              values_fn = mean) %>%
  left_join(road_standards)

comb_score <- comb_stand %>%
  mutate(score_left = getSideScore(standard_sidewalk, standard_shoulder, 
                                   sidewalk_left, shoulder_left),
         score_right = getSideScore(standard_sidewalk, standard_shoulder, 
                                   sidewalk_right, shoulder_right),
         score = score_left + score_right)

@@

comb_score2 <- st_set_geometry(ncdf, NULL) %>% 
  select(RD_LOG_ID_R = RD_LOG_ID, SEGMENT_ID_R = SEGMENT_ID,
         FULL_NAME_R = FULL_NAME, road_class, RUCODE, LENGTH) %>%
  mutate(setting = ifelse(RUCODE == 1, "Rural", 
                          ifelse(RUCODE == 2, "Urban", "Unknown"))) %>%
  left_join(comb_score) %>%
  mutate(the_score = ifelse(is.na(score), 0, score),
         the_length = ifelse(is.na(distance), LENGTH, distance),
         w_score = the_score*the_length)

comb_total_score <- sum(comb_score2$w_score)/sum(comb_score2$the_length)

comb_score_by_setting <- comb_score2 %>%
  group_by(setting) %>%
  summarise(score = sum(w_score)/sum(the_length),
            .groups = "drop") %>%
  rowwise() %>%
  mutate(interpretation = getScoreInterpretation(score))

>>>>

---
title: "Example Kitsap County Pedestrian Facilities Evaluation Report"
date: "`r Sys.Date()`"
bibliography: metrics.bib
output: 
  html_document:
    number_sections: true
    toc: true
    toc_float: true
    toc_depth: 3
    fig_caption: yes
---

@@

@@

library(kableExtra)
library(tidyverse)
library(sf)
source(paste0(this_dir,'functions/reportFunctions.R'))
```

@@

source(paste0(this_dir,'functions/getReportData.R'))

cities <- getDataCities()
county_no_cities <- getDataCountyNoCities()
kitsap <- getMapKitsap()
clean_roadcl <- getDataRoad()
ncdf <- getDataRoadNoCities(clean_roadcl, county_no_cities)
sidewalks_nc <- getDataSidewalks(county_no_cities)
shoulders_nc <- getDataShoulders(county_no_cities)
nc_sw_sh <- getDataFacilities(sidewalks_nc,shoulders_nc)

@@

```{r moredata, cache = TRUE}

comb_score2 <- getCombScore2(ncdf, nc_sw_sh)

comb_total_score <- getAggregateScore(comb_score2)

comb_map <- getCombMapSFFromScoreAndNCDF(ncdf, comb_score2) 
```

<p style="font-family: times, serif; font-size:24pt; font-weight:bold; font-style:italic">
    Executive Summary
</p>

<div style="font-family: times, serif; font-size:14pt;  font-style:italic">
This is an example of the first in a proposed series of regular Unincorporated Kitsap County Pedestrian Facilities Evaluation Reports.

In this example, pedestrian facilities on Unincorporated Kitsap County roads score **`r nice_comma(comb_total_score)` out of 10** which is evaluated as **`r getScoreInterpretation(comb_total_score)`**. This overall score is weighted by length of centerline roadway and the presence and width of road shoulders or sidewalks compared to the Kitsap County Road Standards.

Details and scores by Commissioner Districts as well as other smaller regions of the county are included in this report.

We are hopeful that future investments will improve the quality of pedestrian facilities across the county, especially around community facilities, like schools and libraries, and in areas designated for urban service - in the Urban Growth Areas.
</div>

@@

```{r kitsapmap}
#| fig.align = "center", 
#| fig.width = 12,
#| fig.height = 8, 
#| out.width = '100%',
#| dpi = 96

@@

getKitsapRoadMap(kitsap, clean_roadcl, ncdf) 
```

@@

These are the roads that will be evaluated.
```{r}

getKableRoadMiles(ncdf)
```

@@

```{r}
getKableFacilityMiles(sidewalks_nc, "Sidewalk", "The minimum sidewalk in the Kitsap County Road Standards is 5 ft")
```

```{r}
getKableFacilityMiles(shoulders_nc, "Shoulder","The minimum shoulder in the Kitsap County Road Standards is 3 ft")
```

@@

In order to do this evaluation, we must combine the roads data (which has information on the class and setting of roads) with the sidewalk and shoulder data (with information about the pedestrian facilities). Here, we join based on the geometries of the roads and sidewalks/shoulders.^[We intersect the road data with a buffered geometry of the sidewalk or shoulder and then remove any matching roads that do not have the same road log id.] Because segments are different for roads and sidewalk shoulder data, a single segment of road centerline may match several segments of sidewalk or shoulder data, and less commonly vice versa.

@@

@@

```{r tbl_standards}
# Road Standards Tables 3.3 and 3.4

getKableRoadStandardsGiven()
```

Because of all of the items that are not available in the road data at this time, this draft makes some assumptions to allow for an evaluation. When the information is known, a more accurate evaluation can be made.^[Roads with unknown setting - RUCODE is 0 are evaluated to the Urban road standards. This may be incorrect.]

```{r tbl_standards_simple}
# Road Standards Tables 3.3 and 3.4 translated for data we have
getKableRoadStandards()
```

@@

- 5 if it has a sidewalk or shoulder of adequate size (based on road classification & standards) or a sidewalk is present where a shoulder is the standard
- 3 if the sidewalk or shoulder is present and is minimum AASHTO requirement (like a 4 ft shoulder)
- 2 if the standard is a sidewalk but a shoulder of adequate size (based on road classification and standards) is present
- 1 if a sidewalk or shoulder is present and is narrower than AASHTO 
- 0 if none present or no data

Note that in the absence of ADT data, our interpretation of the AASHTO minimum is limited - we assume it is a 4 foot shoulder for this example. However, AASHTO minimums can be up to 8 feet under certain conditions.

@@

```{r}
getKableExampleAggregateScore()
```

@@

getScoreInterpretationKable()
```

**NOTE TO SPECIAL COMMITTEE: a single segment can never have a score between 8 and 10. And an aggregation of an area could have a score of 9.9 and still show up as Adequate - it only gets a Good at a 10. Should we consider interpreting anything over an 8 as Good?**

@@

There are road segments with ratings of Nothing (0) through Good (10).

A map of the county with roads colored based on their scores shows the distribution of pedestrian facilities. Urban Growth Areas are highlighted.

```{r map_score_overall}
#| fig.align = "center", 
#| fig.width = 10, 
#| fig.height = 8, 
#| out.width = '100%',
#| dpi = 96
#| 
getMapScoreSummaryOverall(kitsap, uga, cities, comb_map)

@@

getKableScoreSummaryForMap(comb_map)
```

### By Setting and Road Class
Looking by the same breakdown that we have for standards - by setting and road class, we can see where the scores are highest and lowest.

```{r tbl_combscorestandard}
getKableScoreSummaryForStandards(comb_score2)
```

### By Setting Alone

The character of an area contributes greatly to whether we would expect pedestrians. More pedestrians are expected in urban settings than rural.

```{r tbl_combscoresetting}
getKableScoreSummaryForSetting(comb_score2)
```

It may be useful to consider that for the unincorporated county, the roads that are designated as in an Urban setting are mostly inside of the Urban Growth Areas - these areas are identified to develop to an urban service standard, like a city. **With the data and methods used in this example report, these areas are not amenable to pedestrians - contrasting what we might desire in our urban environment.**

## Smaller Regions
We can also look at regions and pedestrian generators within the County. For these, we identify the roads and any sidewalks or shoulders.^[In theory, the miles of recorded sidewalk or shoulder and matched miles could be double the centerline miles if every mile had a sidewalk or shoulder.]
